Visualising Life Satisfaction in Countries across Europe: Individual- and Multi-Level Evidence on Cultural and Pecuniary Factors

### 2019-01-31 LaTex/Paper_fulltext.tex
Final ones

	\section{Project Scope and Preliminary Remarks}
	Our project investigates life satisfaction across Europe and puts an emphasis on the data visualisation part that comes
	with communicating research results. This aspect appears most relevant with regard to `Open Science' and `Citizen
	Science' approaches that are gaining importance. In order to cater to this motivation, we developed an interactive interface
	- The European Happiness Observer - to inspect life satisfaction and its various antecedents across Europe. The app offers
	a myriad of possibilities to inspect life satisfaction on a descriptive basis or delve deeper by computing regression and multilevel
	analyses. Here, the user can select various options and combinations from a pre-selected pool of variables. \\
	Consciously, the scope of our visualisation tool is kept wider and should steer curiosity and inclination to develop own 
	happiness-related (research) questions. To assure at least some theoretical embeddedness, this paper elaborates upon the
	relation of life satisfaction and work-related variables. This perspective served as a broad orientation during the development
	phase of our tool, illustrates the variable selection and comprehensively exhibits what kind of aspects happiness research
	can elucidate. At this point it should be stressed that the Observer enables the user to examine the questions raised in this
	paper. However, the tool's scope goes far beyond these applications and invites every user to get creative in her own 
	right. Thus, this background paper fulfils two major functions: One the one hand, it communicates the data-driven
	approach of our visualisations i.e., the European Happiness Observer app. On the other hand, it documents the theoretical foundation
	of our variable selection and exemplifies a major question of happiness research, namely the relation to work-related
	variables.

@@

	\section{Theoretical Background and Methodological Implementation}
	Although the causal relation between work and happiness/life satisfaction seems straightforward and intuitively correct,
	its academic investigation is far from trivial. On the one hand, this appears in light of definitional questions regarding the 
	wo variables of interest. On the other hand, operationalising the constructs appropriately remains subject to discussions
	(for subjective well-being see: \citet{kahneman_developments_2006,layard_measuring_2010}). Lastly, the causal nature of job and life
	satisfaction presents itself as far more complex than intuitively assumed. \\
	Each of these three fields is addressed by a comprehensive literature review to set the tone for the subsequent analysis.
	As this analysis draws on specific independent variables on the individual level and several factors on higher aggregate
	levels (multi-level analysis), their respective relation to life satisfaction, as assessed in the literature, is discussed as well.

@@

	\subsection{SWB and Job Satisfaction}
	The literature examining the causal relation of SWB and job satisfaction is extensive
	(for reviews see: \citet{bowling_meta-analytic_2010,erdogan_whistle_2012}). \citet{bowling_meta-analytic_2010}, in their
	meta-analysis, show a positive relationship between all three respective components of SWB and job satisfaction.
	Going even further, \citet{near_job_1987} postulate that “job satisfaction must logically constitute some part of the
	broader construct of life satisfaction; thus, the question regarding spillover between the two must be ‘how much’ not
	‘whether’." (p. 398). \\
	Before we examine the how much-question, it is necessary to discuss the direction of causality between SWB and job
	satisfaction. In general, the present literature differentiates two approaches: the part-whole theory
	\citep{bowling_meta-analytic_2010,hart_predicting_1999,judge_dispositional_1998} and the dispositional approach. The former
	proposes job satisfaction to act upon SWB. This appears because of the simple reason that has already been mentioned
	above: job satisfaction constitutes a sub-category of overall life satisfaction \citep[cf.]{near_job_1987}.
	Consequently, the approach is also conceived of as bottom-up approach \citep{erdogan_whistle_2012}. Conversely, the
	dispositional approach assumes a general predisposition of satisfaction with life to affect more specific domains like work
	i.e., satisfaction with those domains \citep{bowling_meta-analytic_2010,diener_subjective_1984,headey_top-down_1991,judge_job_1993,judge_another_1993}. 
	Erdogan et al. (2012) link these predispositions to personal traits that determine how satisfied people feel with certain
	domains and life in general (top-down approach). Positioned between the poles is the spillover hypothesis that just
	contents the basic linkage between job and life satisfaction or even other domains of life 
	\citep{bowling_meta-analytic_2010,erdogan_whistle_2012,judge_job_2001,judge_individual_1994}.
	Conceptually, this relation can be understood as a reciprocal relation between the various domains of life and people’s
	satisfaction (Bowling et al., 2010). \\
	While the direction of causality between life satisfaction and job satisfaction is an open discussion, the interplay of life
	satisfaction with other work-related variables might be easier to grasp. Indeed, our study also examines the effects of
	variables that characterise the working environment beyond perceived job satisfaction. This wider scope appears
	reasonable since “job satisfaction is only a partial indicator of satisfaction with work domain. A person’s satisfaction with
	career and work-related stress are important elements of the work domain that also need to be understood.” \citep[p. 1042]{erdogan_whistle_2012}.

	\subsection{The Wider Interplay of Work-Related Variables with Life Satisfaction}
Our analysis features a myriad of variables that can be examined with regard to their effect on overall life satisfaction.
In order to justify our variable selection the following remarks elucidate the respective variables' interplay with life satisfaction.
In contrast to the previous examination of the overall concept of SWB, to accounts on the different variables are presented 
in a briefer manner. Rather than covering the entire theoretical background, the following elaboration is though to spur interest
and curiosity to examine the relation of life satisfaction and other factors on one's own - aided by our interactive 
European Happiness Observer interface.

@@

\section{User Manual}
The section is aimed at enabling the user to use the project results competently. To do so, a working model of the statistical software R is required, the following instructions are made for application within its GUI RStudio. The project is divided in 'Data preparation', "Varaible Preparation', 'Visualisation' and 'app'. The latter is the most important for the first time user as it will provide access to the reactive visualisation tool 'European Happiness Observer' that is the centrepiece of this project. First time users, may ignore the other files and can execute the whole 'app' file after installing all required packages. The visualisation tool will open itself subsequently, no further steps or data provision needed. The application has guidelines itself that should be helpful for the user in navigating through its function. Generally speaking, it offers opportunities to build mapplots for the nations and most of the regions at NUTS-1 level, other visualisations can be made using bar charts, bivariate line plots and boxplots grouped by geographical entities. Additionally, it is possible to adapt some linear regression models and display their output. Finally, raw data tables are displayed as the most unrefined view on the data. While the mapplots, bar charts and line plots can be understood quite intuitively, this paragraph is aimed at giving some guidelines for working with the 'Boxplot' and 'Regression table' elements in the application.\\

A boxplot is a visualisation that displays the distribution of data on the y-axis,  The dark bar inside the box indicates the position of the median, while the box itself indicates the value for the first and third quartile of the data. Thus the inside of the box contains 50\% of the data. The antennas around the box (also called 'whiskers') have a distance from it of 1.5 times the size of the box (distance between the third and first quartile, called interquartile range). Any data points outside the range of box and whiskers are considered outliers from the data distribution. Thus, the boxplot allows to compare the spread of data for different groups that are displayed on the x-axis. In the application it is possible, to order the groups after certain variables to see, whether the variable may have an effect on the spread of data.\\

@@

The regression tables provide the user with the opportunity to compile customized regression models that will give formally statistical evidence on the substance of relations in the data. The simple linear regression without multilevel elements is to be interpreted very similar to the line plot, as starting with the constant value, every increase in the independent variable changes the dependent variable by a certain amount. The number of stars next to the independent variables indicate the significance of its effect (significance as in 'being present'). The total validity of the model can be taken from the value of R$^2$ as it indicates the share of data spread in the dependent variable that is explained by the data spread in the independent variable. In table 1, this is exemplified for the model from 'Job satisfaction' on 'Life satisfaction'.

@@

Users might be interested in seeing more detailed information on variables and can do so in the codebook 'EVS\_final\_cdbk' in the project file. It is also possible to generate the codebook anew by executing the 'Visualisations file'. if users may wish to make changes to data or transformation of variables, the may do so in the files 'Data preparation' and 'Variable preparation'.

\section{Conclusion}
To conclude this project description, it remains to say that the project outline was, at times, torn between several 
different, sometimes conflicting expectations. The ambition to produce high-quality analytical insights combined with 
deep-diving theoretical contextualisation is a challenging endeavour. Conducting a project that allows easily accessible 
data exploration for lay people is tough in its own right. One the one hand, one has to cater to an audience that has only
little understanding of the statistics or programming behing the figures. On the other hand, 
offering enough substance and educational value to spark further interest on the topic and the method should be assured.
These are expectational challenges that are similar to the aforementioned trade-off between 'Open Science' and 'Citizen
Science'. However, the result of this project is promising in so far as it fulfils many of the expectations that it was set out
to meet. The application produced is easily accessible and offers a range of possibilities for the analysis of various
variables on an individuals, regional, and national level. 
Moreover, data sourcing and processing is thoroughly documented. Lastly, significant theoretical considerations
justify our selection of variables and methods and will hopefully attract others to explore the topic of life satisfaction.\\
Future challenges include sufficient data access and quality, especially regarding countries and regions that are not covered to date.
Despite - or maybe because - of future challenges we are optimistic that this project is just the first destination of a longer journey. 
	%% The Appendices part is started with the command \appendix; appendix sections are then done as normal sections
\newpage
 \appendix

>>>>

	\section{Project Scope and Preliminary Remarks}
	Our project investigates life satisfaction across Europe and puts an emphasis on the data visualisation part that comes
	with communicating research results. This aspect appears most relevant with regard to `Open Science' and `Citizen
	Science' approaches that are gaining importance. In order to cater to this motivation, we developed an interactive interface
	- The European Happiness Observer - to inspect life satisfaction and its various antecedents across Europe. The app offers
	a myriad of possibilities to inspect life satisfaction on a descriptive basis or delve deeper by computing regression and multilevel
	analyses. Here, the user can select various options and combinations from a pre-selected pool of variables. \\
	Consciously, the scope of our visualisation tool is kept wide and should steer curiosity and inclination to develop own 
	happiness-related (research) questions. To assure at least some theoretical embeddedness, this paper elaborates upon the
	relation of life satisfaction and work-related variables. This perspective served as a broad orientation during the development
	phase of our tool, illustrates the variable selection and comprehensively exhibits what kind of aspects happiness research
	can elucidate. At this point it should be stressed that the Observer enables the user to examine the questions raised in this
	paper. However, the tool's scope goes far beyond these applications and invites every user to get creative in her own 
	right. Thus, this background paper fulfils two major functions: One the one hand, it communicates the data-driven
	approach of our visualisations i.e., the European Happiness Observer app. On the other hand, it documents the theoretical foundation
	of our variable selection and exemplifies a major question of happiness research, namely the relation to work-related
	variables.

@@

	\section{Theoretical Background and Methodological Implementation}
	Although the causal relation between work and happiness/life satisfaction seems straightforward and intuitively correct,
	its academic investigation is far from trivial. On the one hand, this appears in light of definitional questions regarding the 
	two variables of interest. On the other hand, operationalising the constructs appropriately remains subject to discussions
	(for subjective well-being see: \citet{kahneman_developments_2006,layard_measuring_2010}). Lastly, the causal nature of job and life
	satisfaction presents itself as far more complex than intuitively assumed. \\
	Each of these three fields is addressed by a comprehensive literature review to set the tone for the subsequent analysis.
	As this analysis draws on specific independent variables on the individual level and several factors on higher aggregate
	levels (multi-level analysis), their respective relation to life satisfaction, as assessed in the literature, is discussed as well.

@@

	\subsection{SWB and Job Satisfaction}
	The literature examining the causal relation of SWB and job satisfaction is extensive
	(for reviews see: \citet{bowling_meta-analytic_2010,erdogan_whistle_2012}). \citet{bowling_meta-analytic_2010}, in their
	meta-analysis, show a positive relationship between all three respective components of SWB and job satisfaction.
	Going even further, \citet{near_job_1987} postulate that “job satisfaction must logically constitute some part of the
	broader construct of life satisfaction; thus, the question regarding spillover between the two must be ‘how much’ not
	‘whether’." (p. 398). \\
	Before examining the how much-question, it is necessary to discuss the direction of causality between SWB and job
	satisfaction. In general, the present literature differentiates two approaches: the part-whole theory
	\citep{bowling_meta-analytic_2010,hart_predicting_1999,judge_dispositional_1998} and the dispositional approach. The former
	proposes job satisfaction to act upon SWB. This appears because of the simple reason that has already been mentioned
	above: job satisfaction constitutes a sub-category of overall life satisfaction \citep[cf.]{near_job_1987}.
	Consequently, the approach is also conceived of as bottom-up approach \citep{erdogan_whistle_2012}. Conversely, the
	dispositional approach assumes a general predisposition of satisfaction with life to affect more specific domains like work
	i.e., satisfaction with those domains \citep{bowling_meta-analytic_2010,diener_subjective_1984,headey_top-down_1991,judge_job_1993,judge_another_1993}. 
	Erdogan et al. (2012) link these predispositions to personal traits that determine how satisfied people feel with certain
	domains and life in general (top-down approach). Positioned between the poles is the spillover hypothesis that just
	contents the basic linkage between job and life satisfaction or even other domains of life 
	\citep{bowling_meta-analytic_2010,erdogan_whistle_2012,judge_job_2001,judge_individual_1994}.
	Conceptually, this relation can be understood as a reciprocal relation between the various domains of life and people’s
	satisfaction (Bowling et al., 2010). \\
	While the direction of causality between life satisfaction and job satisfaction is an open discussion, the interplay of life
	satisfaction with other work-related variables might be easier to grasp. Indeed, our study also examines the effects of
	variables that characterise the working environment beyond perceived job satisfaction. This wider scope appears
	reasonable since “job satisfaction is only a partial indicator of satisfaction with work domain. A person’s satisfaction with
	career and work-related stress are important elements of the work domain that also need to be understood.” \citep[p. 1042]{erdogan_whistle_2012}.

	\subsection{The Wider Interplay of Work-Related Variables with Life Satisfaction}
Our analysis features a myriad of variables that can be examined with regard to their effect on overall life satisfaction.
In order to justify our variable selection the following remarks elucidate the respective variables' interplay with life satisfaction.
In contrast to the previous examination of the overall concept of SWB, to accounts on the different variables are presented 
in a briefer manner. Rather than covering the entire theoretical background, the following elaboration is thought to spur interest
and curiosity in examining the relation of life satisfaction and other factors on one's own - aided by our interactive 
European Happiness Observer interface.

@@

\section{User Manual}
The section is aimed at enabling the user to use the project results competently. To do so, a working model of the statistical software R is required, the following instructions are made for application within its GUI RStudio. The project is divided in 'Data preparation', "Variable Preparation', 'Analysis, 'Visualisation' and 'app'. The latter is the most important for the first-time user as it will provide access to the reactive visualisation tool 'European Happiness Observer' that is the centrepiece of this project. First time users may ignore the other files and can execute the whole 'app' file after installing all required packages. The visualisation tool will open itself subsequently, no further steps or data provision needed. The application has guidelines itself that should be helpful for the user in navigating through its functions. Generally speaking, the app offers opportunities to build mapplots for the nations and most of the regions at NUTS-1 level, other visualisations can be made using bar charts, bivariate line plots and boxplots grouped by geographical entities. Additionally, it is possible to adapt some linear and multilevel regression models and display their output. Finally, raw data tables are displayed as the most unrefined view on the data. While the mapplots, bar charts and line plots can be understood quite intuitively, this paragraph is aimed at giving some guidelines for working with the 'Boxplot' and 'Regression table' elements in the application.\\

A boxplot is a visualisation that displays the distribution of data on the y-axis, the dark bar inside the box indicates the position of the median, while the box itself indicates the value for the first and third quartile of the data. Thus the inside of the box contains 50\% of the data. The antennas around the box (also called 'whiskers') have a distance from it of 1.5 times the size of the box (distance between the third and first quartile, called interquartile range). Any data points outside the range of box and whiskers are considered outliers from the data distribution. Thus, the boxplot allows to compare the spread of data for different groups that are displayed on the x-axis. In the application it is possible, to order the groups after certain variables to see, whether the variable may have an effect on the spread of data.\\

@@

The regression tables provide the user with the opportunity to compile customized regression models that will give formal statistical evidence on the substance of relations in the data. The simple linear regression without multilevel elements is to be interpreted very similar to the line plot, as starting with the constant value, every increase in the independent variable changes the dependent variable by a certain amount. The number of stars next to the independent variables indicate the significance of its effect (significance as in 'being present'). The total validity of the model can be taken from the value of R$^2$ as it indicates the share of data spread in the dependent variable that is explained by the data spread in the independent variable. In table 1, this is exemplified for the model from 'Job satisfaction' on 'Life satisfaction'.

@@

Users might be interested in seeing more detailed information on variables and can do so in the codebook 'EVS\_final\_cdbk' in the project file. It is also possible to generate the codebook anew by executing the 'Visualisation' file. If users wish to make changes to data or transformation of variables, the may do so in the files 'Data preparation' and 'Variable preparation'.

\section{Conclusion}
To conclude this project description, it remains to say that the project outline was, at times, torn between several 
different, sometimes conflicting expectations. The ambition to produce high-quality analytical insights combined with 
deep-diving theoretical contextualisation is a challenging endeavour. But also conducting a project that allows easily accessible 
data exploration for lay people is tough in its own right. One the one hand, one has to cater to an audience that has only
little understanding of the statistics or programming behing the figures. On the other hand, 
offering enough substance and educational value to spark further interest on the topic and the method should be assured.
These are expectational challenges that are similar to the aforementioned trade-off between 'Open Science' and 'Citizen
Science'. However, the result of this project is promising in so far as it fulfils many of the expectations that it was set out
to meet. The application produced is easily accessible and offers a range of possibilities for the analysis of various
variables on an individuals, regional, and national level. 
Moreover, data sourcing and processing is thoroughly documented. Lastly, significant theoretical considerations
justify our selection of variables and methods and will hopefully attract others to explore the topic of life satisfaction.\\
Future challenges include sufficient data access and quality, especially regarding countries and regions that are not covered to date.
Despite - or maybe because - of future challenges we are optimistic that this project is just the first destination of a longer journey. 
	%% The Appendices part is started with the command \appendix; appendix sections are then done as normal sections
\newpage
 \appendix
